\vspace{30mm} {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\ {\large Document Version \DuoPLICRev}

### 2023-06-11 doc/src/CSRs.tex

@@

@@

The space of registers selectable by \z{vsiselect} is more limited than
for machine and supervisor levels:
\begin{displayLinesTable}[l@{\quad}l]
\z{0x000}--\z{0x02F} & reserved \\
\z{0x030}--\z{0x03F} & inaccessible \\
\z{0x040}--\z{0x06F} & reserved \\
\z{0x070}--\z{0x0FF} & external interrupts (IMSIC only), or inaccessible \\
\z{0x100}--\z{0x1FF} & reserved \\
\end{displayLinesTable}
When \z{vsiselect} has the number of an \emph{inaccessible} register,
attempts from \mbox{M-mode} or \mbox{HS-mode} to access \z{vsireg}
raise an illegal instruction exception, and attempts from
\mbox{VS-mode} or \mbox{VU-mode} to access \z{sireg}
(really \z{vsireg}) raise a virtual instruction exception.
Similarly, when \z{vsiselect} has a reserved value,
including values above \z{0x1FF} without the most-significant bit set,
attempts from \mbox{M-mode} or \mbox{HS-mode} to access \z{vsireg}
should preferably raise an illegal instruction exception, and attempts from
\mbox{VS-mode} or \mbox{VU-mode} to access \z{sireg} (really \z{vsireg})
should preferably raise a virtual instruction exception.

@@

The locations of registers for external interrupts (numbers
\z{0x70}--\z{0xFF}) are accessible only when field VGEIN of \z{hstatus}
is the number of an implemented guest external interrupt, not zero.
If VGEIN is not the number of an implemented guest external interrupt
(including the case when no IMSIC is implemented), then all non-custom
values of \z{vsiselect} are either reserved
or designate an inaccessible register.

@@

@@

If one of these bits is zero in \z{hstateen0},
and the same bit is one in \z{mstateen0},
then an attempt to access the corresponding state
from VS or \mbox{VU-mode} raises a virtual instruction exception.
(But note that, for high-half CSRs \z{siph} and
\z{sieh}, this applies only when XLEN =~32.
When $\mbox{XLEN} > \mbox{32}$, an attempt to access
\z{siph} or \z{sieh} raises an illegal instruction
exception as usual, not a virtual instruction exception.)

>>>>

If extension Smcsrind if also implemented, then when \z{miselect}
has a value in the range \z{0x30}--\z{0x3F} or \z{0x70}--\z{0xFF},
attempts to access alias CSRs \z{mireg2} through \z{mireg6}
raise an illegal instruction exception.

@@

If extension Sscsrind if also implemented, then when \z{siselect}
has a value in the range \z{0x30}--\z{0x3F} or \z{0x70}--\z{0xFF},
attempts to access alias CSRs \z{sireg2} through \z{sireg6}
raise an illegal instruction exception (unless executing
in a virtual machine, covered in the next section).

@@

The space of registers selectable by \z{vsiselect} is more limited than
for machine and supervisor levels:
\begin{displayLinesTable}[l@{\quad}l]
\z{0x000}--\z{0x02F} & reserved \\
\z{0x030}--\z{0x03F} & inaccessible \\
\z{0x040}--\z{0x06F} & reserved \\
\z{0x070}--\z{0x0FF} & external interrupts (IMSIC only), or inaccessible \\
\z{0x100}--\z{0x1FF} & reserved \\
\end{displayLinesTable}

When \z{vsiselect} has a reserved value (including values above
\z{0x1FF} not designated for custom use), attempts from \mbox{M-mode}
or \mbox{HS-mode} to access \z{vsireg}, or from \mbox{VS-mode}
or \mbox{VU-mode} to access \z{sireg} (really \z{vsireg}),
should preferably raise an illegal instruction exception.

When \z{vsiselect} has the number of an \emph{inaccessible} register,
attempts from \mbox{M-mode} or \mbox{HS-mode} to access \z{vsireg}
raise an illegal instruction exception, and attempts from
\mbox{VS-mode} or \mbox{VU-mode} to access \z{sireg}
(really \z{vsireg}) raise a virtual instruction exception.

@@

The locations of registers for external interrupts (numbers
\z{0x70}--\z{0xFF}) are accessible only when field VGEIN of \z{hstatus}
is the number of an implemented guest external interrupt, not zero.
If VGEIN is not the number of an implemented guest external interrupt
(including the case when no IMSIC is implemented), then all
indirect register numbers in the ranges \z{0x030}--\z{0x03F} and
\z{0x070}--\z{0x0FF} designate an inaccessible register at VS level.

@@

If extension Sscsrind if also implemented, then when \z{vsiselect} has
a value in the range \z{0x30}--\z{0x3F} or \z{0x70}--\z{0xFF},
attempts from privilege modes other than VS or VU to access alias CSRs
\z{vsireg2} through \z{vsireg6} raise an illegal instruction exception,
and attempts from VS or \mbox{VU-mode} to access \z{sireg2}
through \z{sireg6} or \z{vsireg2} through \z{vsireg6}
raise a virtual instruction exception.

@@

If one of these bits is zero in \z{hstateen0},
and the same bit is one in \z{mstateen0},
then an attempt to access the corresponding state
from VS or \mbox{VU-mode} raises a virtual instruction exception,
unless one of the following cases applies
to cause an illegal instruction exception instead:
\begin{itemize}

\item
An attempt to access \z{sireg} raises an illegal instruction exception
if the same attempt would do so when bit~60 of \z{hstateen0} is one.

\item
When $\mbox{XLEN} > \mbox{32}$, attempts to access high-half CSRs
\z{siph} and \z{sieh} always raise an illegal instruction exception.

\end{itemize}
